Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-09-10 _manuscript/Manuscript.qmd
Manuscript rendering

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
bibliography: references.bib
---

@@

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities. The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability. At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

This description of the DOM model form and sampling design are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003. Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

DOMs make a number of assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science: the first of these included all articles from 2004-2023 which cite @mackenzie2003; the second query searched articles in the same time-span matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*", articles including each of "occupancy", "colonization", "extinction/persistence", and "detection", and articles with the term "occupancy" located near "dynamic" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided these articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

@@

Many articles in our sample used the DOM and sought to monitor trends in occupancy state through time. These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle. These studies offer valuable insights on the state of species across landscapes and through time, and DOMs are a strong candidate for many monitoring oriented projects — including those where many species need to be assessed at once. Other articles were more focused on examining the factors which influence species occupancy, either testing a pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception. The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models. Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl. Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions as in @miller2023's study on American Kestrel nest boxes. In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as illustrated in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy. In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy. As examples, @mcgowan2020 provides projections for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.

@@

These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species @carvalho2023, guiding public health management of disease vectors @moreira2016, and tracking rapidly developing biological invasions @wood2020. DOMs are well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important. In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied. Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM format more accessible to newcomers and support confidence in interpretation of models. Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] while specifically addressing model attributes unique to DOMs.

>>>>

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.

bibliography: references.bib
csl: apa.csl
---

@@

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities. The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability. At sites which are truly unoccupied, it is assumed that the species is never detected (i.e., no false positives).](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

DOMs make a number of key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, model extensions designed to account for false positive error should be considered.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To capture any additiona relevant articles which did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

@@

Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time. These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle. These studies offer valuable insights on the state of species across landscapes and through time, demonstrating DOMs' suitability for many monitoring oriented projects — including those where many species need to be assessed at once. Other articles were more focused on examining the factors which influence species occupancy, either testing a pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception. The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models. Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl. Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions as in @miller2023's study on American Kestrel nest boxes. In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as illustrated in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy. In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy. As examples, @mcgowan2020 provides projections for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.

@@

These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species [@carvalho2023], guiding public health management of disease vectors [@moreira2016], and tracking rapidly developing biological invasions [@wood2020]. DOMs are well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important. In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied. Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM format more accessible to newcomers and support confidence in interpretation of models. Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims [@zurell2020; @araújo2019] while specifically addressing model attributes unique to DOMs.
